Towards a Wide-Coverage Grammar for Swedish\\ Using GF

### 2012-01-30 doc/report/report.tex
cover and pics to report

\begin{document}
% ska gå att förstå och locka till läsning
%ska tydligt beskriva vad rapporten handlar om
%ska innehålla vettiga sökord
\title{Towards a Wide-Coverage Grammar for Swedish\\ Using GF} %A nice titleScaling up a GF grammar and preparing it for cool stuff for Swedish}
\author{Malin Ahlberg}
\maketitle
\newpage

@@

%> dependent and implicit types : agda. (read gf book and krasses book)
A grammar written in GF can be used for both parsing and generation.
The parsing algorithm is incremental and has polynomial time and space
complexity \cite{gfMech}. 
The GF package also provides various tools for working with and using the grammars:
a compiler, an interpreter and a runtime system.
The grammars can be compiled into portable grammar format (PGF) \cite{pgf},
supported by %There are libraries for using PGF files with
Java, Java script, C and Haskell libraries. 
%>> Haskell only complete, also or C\# and C.  %may also be used in Android applications.
The interpreter, the GF shell, allows
the user to test grammars by commands for parsing, visualization of parse trees,
random generation, word alignment, morphological quizzes etc.
The shell can be tried out online 
together with an interactive GF editor %and the GF shell 
\footnote{http://www.grammaticalframework.org/demos/gfse/}.
\thispagestyle{plain}
\begin{figure}[h]
\begin{verbatim} 
> parse "jag ser katten" | visualize_tree  -format=pdf -view=evince
> parse "jag ser katten" | visualize_parse -format=pdf -view=evince
\end{verbatim}
\caption{Example of how the GF shell can be used.}\label{fig:shellvp}
\label{fig:shell}
\end{figure}
Figure \ref{fig:shell} shows commands for parsing, where the results are shown as
in figure \ref{fig:gftree1}. 
GF uses Graphviz\footnote{http://www.graphviz.org/}
to visualize the trees and the given commands (fig. \ref{fig:shell}) specify that the
output format should be PDF and that these files should be opened by the program
Evince. The user can choose to see the \textit{parse tree} (fig. \ref{pic:gfStree})
or the \textit{abstract tree} (fig. \ref{pic:gfAtree}). Abstract trees are more verbose and show all
functions and types used for parsing the sentence. 
\begin{figure}[h]
\centering
\subfloat[Swedish parse tree]{\label{pic:gfStree}\includegraphics[width=30mm]{gfSTree.png}}
\subfloat[GF abstract tree]{\label{pic:gfAtree}\includegraphics[width=100mm]{gfTree.pdf}}
\caption{Abstract tree and parse tree for the sentence ``Jag ser katten".}
\label{fig:gftree1}
\end{figure}
\vspace{10mm}

@@

%>>the scope is controlled language, but possible ambiguities (discuss this in future
%>>work, saldo, conclusion). Ambiguities: from multilinguality, or inherited:
%>>syntactical (man with the telescope), lexical (youSg/Pl, växa).
%>> have ambiguity section. 
\newpage
\begin{wrapfigure}{l}{0.35\textwidth}
\includegraphics[width=30mm]{gfETree.png}
\caption{English parse tree}
\label{pic:gfEtree}
\end{wrapfigure}

@@

\section{Saldo}
\label{sec:saldo}
A good parser needs a good lexicon. We have used Saldo \cite{saldo}, a
large electronic lexicon developed and maintained at Gothenburg University. It is
built on Svenskt Associationslexikon and contains information about more than 
120 000 modern Swedish words.
For each word there is semantic, syntactical and a morphological
information. The user can find examples of usage in corpora,
graphs of semantically connected words and some suggestions for how to analyse compounds. \\
\begin{figure}[h!]
\centering
\subfloat[Morphological info for `katt']{\label{pic:saldoTab}\includegraphics[width=40mm]{saldotab.png}}
\hspace{20mm}
\subfloat[Graph showing the hyponyms of `katt']{\label{pic:saldoWheel}\includegraphics[width=70mm]{saldograph.png}}
\caption{Information in Saldo}
\label{fig:saldo}
\end{figure}

@@

Figure \ref{pic:TabVax} shows the information given by Saldo and by GF respectively.\\
\begin{figure}[h]
  \begin{center}
\subfloat[Saldo]{\label{pic:saldoVax}\includegraphics[height=90mm]{saldoVax.png}}
\hspace{5mm}
\subfloat[GF]{\label{pic:gfVax}\includegraphics[height=60mm]{gfVax.png}}
\caption{Inflection table for the verb `växa' (grow)}
\label{pic:TabVax}
  \end{center}
\end{figure}
The tables does not overlap completely, there are some more forms in Saldo's table (e.g. the 
compounding form `väx-')
while the one generated in GF contains some that are not in Saldo (e.g. `vuxits').
As GF concerns about syntax only, and not semantics, and  as the GF tables are automatically
generated, they always contain all word forms, although some forms may never be
used in the natural language.
Saldo may also contain variants: `växt' and `vuxit' are both supine form.
As far as possible, the program makes up for this by only comparing the overlapping forms
and only requiring that GF generates one variant whenever alternatives are given. \\
%The program continues to generate new lexicons as long as there are more
%paradigms to choose from and compare the results after each run.

@@

The GF resource grammars gives a fundamental description of Swedish,
covering the morphology,
word order, agreement, tense, basic conjunction etc.
 %and many phrase and clause constructions.
Due to the syntactic similarities of the Scandinavian languages, much of the implementation
is shared with Norwegian and Danish. The modules that concern the lexical
aspects are separate, while 85 \% of the syntax description is shared.
There are about 80 functions, which describe the rules for building phrases and clauses.
\begin{verbatim}
PredVP       : NP -> VP -> Cl ; -- Predication
ComplVPSlash : VPSlash  -> VP ; -- Complementation
\end{verbatim}
The analysis preformed by GF is driven by the parts of speech, which are combined into
parts of sentences.
Figure \ref{fig:gfpic} shows the different categories, or types, used in the
resource grammars. 
%A lexicon normally consists of 
Words from the open word-classes are shown in rectangular boxes in
the picture. Each lexical entry is assigned a type describing its word-class.
%and the abstract grammar can be viewed as a set rules showing
%how the parts of speech can be used --
%how the types relate to each other.
\begin{figure}[h]
\hspace{-18mm}
\includegraphics[height=130mm]{categories.png}
\caption{The types of the GF resource grammars. }
%TODO have footnote? where?
%\footnote{Available at http://www.grammaticalframework.org/lib/doc/synopsis.html}
\label{fig:gfpic}
\end{figure}

@@

\noindent Swedish verbs may take up to five arguments \cite[p. 53]{stymne}.
These may be prepositions, particles, reflexive object, indirect objects
and direct objects. 
\enumsentence{\shortex{7}
{Jag &tar& med &mig& den& till& honom}
{\emph{I} &\emph{take}&\emph{with} &\emph{me}& \emph{it}& \emph{to}& \emph{him}}
{``I bring it to him"}
\label{sent:tamed}}
The verb `ta' in sentence (\ref{sent:tamed}) takes one particle (\emph{`med'}),
one preposition (\emph{`till'}) and two objects: \emph{`den'} and \emph{`honom'}.
In a GF lexicon this verb is given the
category \verb-V3-, %classifing it as 
a \textit{three-place verb}, taking two objects.
The notion \verb-V3- is motivated by the formal translation: \verb-bring(I,it,him)-.\\
%these will be encoded in different ways.
Particles are given in the lexicon, as well as the prepositions that are
chosen by the verb, since these may not vary.
The entry for `\emph{ta med}', as used in sentence (\ref{sent:tamed}), is
described as follows:
%, assuming that we already have an constant \verb-take_V- which produces the correct conjugation for `ta':
\begin{verbatim}
ta_med_V3 = dirV3 (reflV (partV (take_V "med"))) (mkPrep "till") ;
\end{verbatim}
%The category \verb-V3- tells us that the verb takes two objects, % (`den' and `honom').
The function \verb-dirV3- creates a three-place verb, where the first object is direct and
the second is to be used with the preposition given as the last argument: `\emph{till}'.
\verb-reflV- shows that the verb always is used with a reflexive
pronoun and \verb-partV- gives the particle \emph{`med'}.
The fact that the chosen prepositions is attached to the verb in the lexicon
causes the parse tree visualization algorithm to group them together.
This is also the case for particles, cf. parse tree \ref{pic:tittapart}
and \ref{pic:tittapa}.
\begin{figure}[h]
\subfloat[Verb with a chosen preposition]{\label{pic:tittapa}\includegraphics[height=70mm]{tittapa.jpg}}
\hspace{10mm}
\subfloat[Verb with particle]{\label{pic:tittapart}\includegraphics[height=76mm]{tittapart.pdf}}
\caption{The visualized parse trees do not show the internal difference of chosen prepositions and particles}
\label{fig:translationtrees}
\end{figure}
\newpage
\noindent As already stated, the visualized parse trees is not a complete representation,
even if the verb phrases in the visualizations look the same, the two cases
are treated and represented differently internally. The fronting of the
preposition, as in of sentence (\ref{sent:tittapa}a)., is accepted but
fronting of particles, as in \ref{sent:tittapa}b., is not.
%TODO use png to get correct size
\eenumsentence{
\item[a]\shortex{4}
{På &pojken& tittar& du.}
{\emph{on} &\emph{the boy}& \emph{look}& \emph{you}}
{``You look at the boy."}
\item[b]\shortexnt{6}
{*På& när& du& springer& tittar& jag}
{\emph{~on}& \emph{when}& \emph{you}& \emph{run}& \emph{watch}& \emph{I}}
\label{sent:tittapa}}

@@

In addition to the core resource grammars, which is shared with the other languages implemented in the
library, there is also an extra module, simple called the \verb|Extra| module, where
%In addition to the basic resource grammar,
 language specific constructions may bee added
%>> core is shared but particular lang. has particular features.
%to the module \verb|Extra|.
The functions given here do not have to be translatable
to all other language, but are meant to cover language specific constructions.
Among those were functions for topicalisation (\ref{gfSwe:apple}), %$ although not all
%of them were implemented for Swedish.
%word order, agreement, tempus, adverbs, adjectivs...
%and relatives
%Also info in the Extra module, where more language specific constructions
%can be found.
%For example, \verb|FocObj| fronted the object as in sentence (\ref{gfSwe:apple}).
\enumsentence{\shortex{6}
{Det där & äpplet & vill & jag & inte & ha}
{that & apple & want & I & not &have}
{``I don't want that apple"}}\label{gfSwe:apple}
and for preposition stranding:
\enumsentence{
~~Stranded preposition \hspace{45mm} cf.\\
\begin{tabular}{ll}
{\begin{tabular}{llllll}
Vem& måste& jag& akta& mig &\textbf{för}? \\
\emph{who}& \emph{must}& \emph{I}& \emph{watch out}& \emph{me} &\textbf{\emph{for}}? \\
\end{tabular}}&
{\begin{tabular}{llllll}
\textbf{För}& vem& måste& jag& akta& mig?\\
\emph{\textbf}{for}& \emph{who}& \emph{must}& \emph{I}& \emph{watch out}& \emph{me}?\\
\end{tabular}}\\
~~`Who do I need to watch out \textbf{for}?' &~ `\textbf{For} whom do I need to watch out?'\\
\end{tabular}}

@@

\subsubsection{Parallel parsing/Chunk parsing}
Instead of parsing a whole sentence in one go, we intend
parse parts of it in parallel and then try to combine the pieces, building
the sentence bottom up.
This method should be faster then normal parsing and would
at the same time give robustness;
components that cannot be combined by any grammar rule are combined by a 
meta function. When an unknown construction is encountered, the parser
will not fail but give a partial parse tree with the analysis for the rest of
the sentence \cite{gfMech}.
The output given to the user will be more valuable, since partial parse
trees are more interesting the no parse trees.
%We might end up with partial parse trees, but we would
%be able to parse at least parts of every sentence and give a more
%interesting result to the user.

>>>>

\begin{document}
% ska gå att förstå och locka till läsning
%ska tydligt beskriva vad rapporten handlar om
%ska innehålla vettiga sökord
\title{Towards a Wide-Coverage Grammar for Swedish\\ Using GF} %A nice titleScaling up a GF grammar and preparing it for cool stuff for Swedish}
\author{Malin Ahlberg}
%\maketitle
\newpage

@@

%> dependent and implicit types : agda. (read gf book and krasses book)
A grammar written in GF can be used for both parsing and generation.
The parsing algorithm is incremental and has polynomial time and space
complexity \cite{gfMech}. 
The GF package also provides various tools for working with and using the grammars:
a compiler, an interpreter and a runtime system.
The grammars can be compiled into portable grammar format (PGF) \cite{pgf},
supported by %There are libraries for using PGF files with
Java, Java script, C and Haskell libraries. 
%>> Haskell only complete, also or C\# and C.  %may also be used in Android applications.
The interpreter, the GF shell, allows
the user to test grammars by commands for parsing, visualization of parse trees,
random generation, word alignment, morphological quizzes etc.
The shell can be tried out online 
together with an interactive GF editor %and the GF shell 
\footnote{http://www.grammaticalframework.org/demos/gfse/}.
\thispagestyle{plain}
\begin{figure}[h]
\begin{verbatim} 
> parse "jag ser katten" | visualize_tree  -format=pdf -view=evince
> parse "jag ser katten" | visualize_parse -format=pdf -view=evince
\end{verbatim}
\caption{Example of how the GF shell can be used.}\label{fig:shellvp}
\label{fig:shell}
\end{figure}
Figure \ref{fig:shell} shows commands for parsing, where the results are shown as
in figure \ref{fig:gftree1}. 
GF uses Graphviz\footnote{http://www.graphviz.org/}
to visualize the trees and the given commands (fig. \ref{fig:shell}) specify that the
output format should be PDF and that these files should be opened by the program
Evince. The user can choose to see the \textit{parse tree} (fig. \ref{pic:gfStree})
or the \textit{abstract tree} (fig. \ref{pic:gfAtree}). Abstract trees are more verbose and show all
functions and types used for parsing the sentence. 
\begin{figure}[h]
\centering
\subfloat[Swedish parse tree]{\label{pic:gfStree}\includegraphics[width=30mm]{gfSTreec.pdf}}
\subfloat[GF abstract tree]{\label{pic:gfAtree}\includegraphics[width=100mm]{gfTree.pdf}}
\caption{Abstract tree and parse tree for the sentence ``Jag ser katten".}
\label{fig:gftree1}
\end{figure}
%\vspace{10mm}

@@

%>>the scope is controlled language, but possible ambiguities (discuss this in future
%>>work, saldo, conclusion). Ambiguities: from multilinguality, or inherited:
%>>syntactical (man with the telescope), lexical (youSg/Pl, växa).
%>> have ambiguity section. 
\newpage
\begin{wrapfigure}{l}{0.35\textwidth}
\includegraphics[width=30mm]{gfETreec.pdf}
\caption{English parse tree}
\label{pic:gfEtree}
\end{wrapfigure}

@@

\section{Saldo}
\label{sec:saldo}
A good parser needs a good lexicon. We have used Saldo \cite{saldo}, a
large electronic lexicon developed and maintained at Gothenburg University. It is
built on Svenskt Associationslexikon and contains information about more than 
120 000 modern Swedish words.
For each word there is semantic, syntactical and a morphological
information. The user can find examples of usage in corpora,
graphs of semantically connected words and some suggestions for how to analyse compounds. \\
\begin{figure}[h!]
\centering
\subfloat[Morphological info for `katt']{\label{pic:saldoTab}\includegraphics[width=40mm]{saldoTab.png}}
\hspace{20mm}
\subfloat[Graph showing the hyponyms of `katt']{\label{pic:saldoWheel}\includegraphics[width=70mm]{saldoGraph.png}}
\caption{Information in Saldo}
\label{fig:saldo}
\end{figure}

@@

Figure \ref{pic:TabVax} shows the information given by Saldo and by GF respectively.\\
\begin{figure}[h]
  \begin{center}
\subfloat[Saldo]{\label{pic:saldoVax}\includegraphics[height=80mm]{saldovax.png}}
\hspace{5mm}
\subfloat[GF]{\label{pic:gfVax}\includegraphics[height=60mm]{gfvax.png}}
\caption{Inflection table for the verb `växa' (grow)}
\label{pic:TabVax}
  \end{center}
\end{figure}
The tables does not overlap completely, there are some more forms in Saldo's table (e.g. the 
compounding form `väx-')
while the one generated in GF contains some that are not in Saldo (e.g. `vuxits').
As GF concerns about syntax only, and not semantics, and  as the GF tables are automatically
generated, they always contain all word forms, although some forms may never be
used in the natural language.
Saldo may also contain variants: `växt' and `vuxit' are both supine form.
As far as possible, the program makes up for this by only comparing the overlapping forms
and only requiring that GF generates one variant whenever alternatives are given. \\
%The program continues to generate new lexicons as long as there are more
%paradigms to choose from and compare the results after each run.

@@

The GF resource grammars gives a fundamental description of Swedish,
covering the morphology,
word order, agreement, tense, basic conjunction etc.
 %and many phrase and clause constructions.
Due to the syntactic similarities of the Scandinavian languages, much of the implementation
is shared with Norwegian and Danish. The modules that concern the lexical
aspects are separate, while 85 \% of the syntax description is shared.
There are about 80 functions, which describe the rules for building phrases and clauses.
\begin{verbatim}
PredVP       : NP -> VP -> Cl ; -- Predication
ComplVPSlash : VPSlash  -> VP ; -- Complementation
\end{verbatim}
The analysis preformed by GF is driven by the parts of speech, which are combined into
parts of sentences.
Figure \ref{fig:gfpic} shows the different categories, or types, used in the
resource grammars. 
%A lexicon normally consists of 
Words from the open word-classes are shown in rectangular boxes in
the picture. Each lexical entry is assigned a type describing its word-class.
%and the abstract grammar can be viewed as a set rules showing
%how the parts of speech can be used --
%how the types relate to each other.
\begin{figure}[h]
\hspace{-18mm}
\includegraphics[height=130mm]{Categories.png}
\caption{The types of the GF resource grammars. }
%TODO have footnote? where?
%\footnote{Available at http://www.grammaticalframework.org/lib/doc/synopsis.html}
\label{fig:gfpic}
\end{figure}

@@

\noindent Swedish verbs may take up to five arguments \cite[p. 53]{stymne}.
These may be prepositions, particles, reflexive object, indirect objects
and direct objects. 
\enumsentence{\shortex{7}
{Jag &tar& med &mig& den& till& honom}
{\emph{I} &\emph{take}&\emph{with} &\emph{me}& \emph{it}& \emph{to}& \emph{him}}
{``I bring it to him"}
\label{sent:tamed}}
The verb `ta' in sentence (\ref{sent:tamed}) takes one particle (\emph{`med'}),
one preposition (\emph{`till'}) and two objects: \emph{`den'} and \emph{`honom'}.
In a GF lexicon this verb is given the
category \verb-V3-, %classifing it as 
a \textit{three-place verb}, taking two objects.
The notion \verb-V3- is motivated by the formal translation: \verb-bring(I,it,him)-.\\
%these will be encoded in different ways.
Particles are given in the lexicon, as well as the prepositions that are
chosen by the verb, since these may not vary.
The entry for `\emph{ta med}', as used in sentence (\ref{sent:tamed}), is
described as follows:
%, assuming that we already have an constant \verb-take_V- which produces the correct conjugation for `ta':
\begin{verbatim}
ta_med_V3 = dirV3 (reflV (partV (take_V "med"))) (mkPrep "till") ;
\end{verbatim}
%The category \verb-V3- tells us that the verb takes two objects, % (`den' and `honom').
The function \verb-dirV3- creates a three-place verb, where the first object is direct and
the second is to be used with the preposition given as the last argument: `\emph{till}'.
\verb-reflV- shows that the verb always is used with a reflexive
pronoun and \verb-partV- gives the particle \emph{`med'}.
The fact that the chosen prepositions is attached to the verb in the lexicon
causes the parse tree visualization algorithm to group them together.
This is also the case for particles, cf. parse tree \ref{pic:tittapart}
and \ref{pic:tittapa}.
\begin{figure}[h]
\subfloat[Verb with a chosen preposition]{\label{pic:tittapa}\includegraphics[height=70mm]{tittapac.pdf}}
\hspace{10mm}
\subfloat[Verb with particle]{\label{pic:tittapart}\includegraphics[height=70mm]{tittapartc.pdf}}
\caption{The visualized parse trees do not show the internal difference of chosen prepositions and particles}
\label{fig:translationtrees}
\end{figure}
\newpage
\noindent As already stated, the visualized parse trees is not a complete representation,
even if the verb phrases in the visualizations look the same, the two cases
are treated and represented differently internally. The fronting of the
preposition, as in of sentence (\ref{sent:tittapa}a)., is accepted but
fronting of particles, as in \ref{sent:tittapa}b., is not.
%TODO use png to get correct size
\eenumsentence{
\item[a]\shortex{4}
{På &pojken& tittar& du.}
{\emph{on} &\emph{the boy}& \emph{look}& \emph{you}}
{``You look at the boy."}
\item[b]\shortexnt{6}
{*På& när& du& springer& tittar& jag}
{\emph{~on}& \emph{when}& \emph{you}& \emph{run}& \emph{watch}& \emph{I}}
\label{sent:tittapa}}

@@

In addition to the core resource grammars, which is shared with the other languages implemented in the
library, there is also an extra module, simple called the \verb|Extra| module, where
%In addition to the basic resource grammar,
 language specific constructions may bee added
%>> core is shared but particular lang. has particular features.
%to the module \verb|Extra|.
The functions given here do not have to be translatable
to all other language, but are meant to cover language specific constructions.
Among those were functions for topicalisation (\ref{gfSwe:apple}), %$ although not all
%of them were implemented for Swedish.
%word order, agreement, tempus, adverbs, adjectivs...
%and relatives
%Also info in the Extra module, where more language specific constructions
%can be found.
%For example, \verb|FocObj| fronted the object as in sentence (\ref{gfSwe:apple}).
\enumsentence{\shortex{6}
{Det där & äpplet & vill & jag & inte & ha}
{that & apple & want & I & not &have}
{``I don't want that apple"}}\label{gfSwe:apple}
and for preposition stranding:
\eenumsentence{
\item
Stranded preposition \\ %\hspace{45mm} cf.\\
%\begin{tabular}{ll}
{\begin{tabular}{llllll}
Vem& måste& jag& akta& mig &\textbf{för}? \\
\emph{who}& \emph{must}& \emph{I}& \emph{watch out}& \emph{me} &\textbf{\emph{for}}? \\
\end{tabular}}
\item
{cf.\\
\begin{tabular}{llllll}
\textbf{För}& vem& måste& jag& akta& mig?\\
\emph{\textbf}{for}& \emph{who}& \emph{must}& \emph{I}& \emph{watch out}& \emph{me}?\\
\end{tabular}}\\
~~`Who do I need to watch out \textbf{for}?' &~ `\textbf{For} whom do I need to watch out?'\\
%\end{tabular}}
}

@@

\subsubsection{Parallel parsing/Chunk parsing}
Instead of parsing a whole sentence in one go, we intend
parse parts of it in parallel and then try to combine the pieces into
a parse tree for the whole sentence.
This method should be faster then normal parsing and would
at the same time give robustness;
components that cannot be combined by any grammar rule are combined by a 
meta function. When an unknown construction is encountered, the parser
will not fail but give a partial parse tree with the analysis for the rest of
the sentence \cite{gfMech}.
The output given to the user will be more valuable, since partial parse
trees are more interesting the no parse trees.
%We might end up with partial parse trees, but we would
%be able to parse at least parts of every sentence and give a more
%interesting result to the user.
